\chapter{Introduction}
\label{chap:intro}

\section{Overview}

There are two main approaches of type checking: static and dynamic. Advocates of
first approach say that benefits of static type checking include earlier
detection of bugs and mistakes (e.g. preventing adding string to an integer),
inline documentation and additional data for advanced code competition, useful
information for compiler, which helps to make optimizations (e.g. replacing
virtual calls by direct calls, when type of the caller are known), improved
runtime efficiency (e.g. not necessary to do dynamic dispatching).

On the other hand, there are also advocates of dynamic type checking who states
that languages with such approach have following advantages: interactive dynamic
development (e.g. REPL driven development + TDD, where tests can be rerun
without recompilation of the whole project), fast prototyping (e.g. type
decisions can be postponed, language are much more expressive and more suitable
for development of systems with rapidly changing or unknown requirements).

Static type checking is a good tool, but it necessary to understand that it
doesn't guarantee programs to be bug-free and correct because it provides
compile-time abstraction over run-time behavior of the software system this
means that some errors can be caught, but execution of the program can still go
wrong. Another, more extensive explanation provided by David Maclver in his work
\cite{staticbroken}. Moreover static typing force to make decisions earlier,
slowdown compilation and make interactive development nearly impossible.

Nowadays many people/teams/companies interested in making data-intensive
systems, just take a look at current trends: data-driven approaches, BigData,
machine learning becomes more and more popular. The reason is clear: huge amount
of data allows to extract useful knowledge, but we need some tools to do it.
Probably dynamism is one of the most important thing for such kind of software
systems because majority of data is not fully structured, according to research
project report of Berkeley University \cite{lymanmuch} around 95\%. In cases when
structure is known up front after few steps people or algorithms generates
queries based on run-time information and data also gets dynamic nature.

Dynamic type checking is a good tool for data-intensive or other kind of dynamic
systems, but it also necessary to understand that it has some trade-offs. It is
much harder to refactor, develop and maintain systems with such type checking
because any even small change can break the system and nobody will know it until
fault will occur later in run-time. There are some techniques, which helps to
deal with it like Test-driven development \cite{beck2003test} or Contract-driven
development \cite{meyer2007contract}, but they doesn't solve all the problems.
Sometimes it is pretty handy to define shape of the date up front to check if
current data conform the specification or to generate data samples from such
specifications.

Information about when and what type checking better to use can be found in
\cite{meijer2004static}, it states that use static typing where possible and
dynamic when needed, but what if there is no possibility to choose type checking
system and it is necessary to use dynamically typed language? Can benefits of
static type checking be achieved in such environment?


\section{Objectives}
In this thesis introduced few ways to get in dynamically typed language
alternatives of benefits, which provides static type checking. As a basis
\textbf{Clojure} programming language was taken, but most of the work should be
suitable for languages, which have same or similar properties described in Chapter
\ref{chap:background}.

Initially, described benefits of static type checking and which of them are more
in demand. After that explained ideas, which help to get similar benefits in
dynamically typed language. Also, some of the existing tools and approaches,
which implements this ideas are explained.

More precisely, following points covered:
\begin{itemize}
\item Improved developer experience
\item Better communication
\item More robust software
\end{itemize}

Improved developer experience means that it is static type system in some
languages provides good and clean error messages and it is good to have tools,
which helps to understand a problem and find its location quicker and with less
effort. Moreover, it is important for hosted language to have tool, which
provides consistent error messages across platforms instead of native stack
traces.

Better communication means that dynamically typed languages has a lack of type
annotations and it is hard to understand what functions consumes as parameters
and what produces. As mentioned in \cite{janes2014lean} documentation often
became outdated and useless, that is why solution must be part of codebase for
more leverage and power and be up-to-date (actual for current codebase).

More robust software means that static type checking provides capabilities for
static analysis, which helps to check kind of soundness between components.
In dynamically typed languages it often hard to make static analysis of the
system, but with gradual typing it is possible \cite{tobin2008design}. Another
important tool for creation of robust software is tests and with type
annotations it is possible to generate bunch of them automatically.

This thesis work provides explanation how to achieve points mentioned above
using some kind of optional annotations without breaking existing workflows
(REPL-driven or/and test-driven development approaches), but it is important to
understand that it is always necessary to do trade offs for each solution, that
is why in Chapter \ref{chap:evaluation} provided an explanation of the cost of
such solution.
% On top of that, the example of the tool, which shows how to automate the process
% of generating tests and enriching documentation string in run-time is introduced
% as a part of contribution of this thesis. This tool provides an interface
% (functions), which allows to modify meta-data of functions in run-time
% namespace-wide or project-wide and another interface (macros), which allows to
% generate test for annotated function in a single namespace.



% \section{Outline}
% Chapter \ref{chap:background} provides an overview of benefits, existing tools
% and possible alternatives






% % \section{Draft section}
% This is the introductory chapter.  This will give you some
% % ideas on how to use \LaTeX~\cite{lam1994} to typeset your document.
% Here is a sample quote using the \verb+\munquote+ environment:

% % \begin{munquote}[~\cite{lam1994}]%
% % \LaTeX{} is a system for typesetting documents.  Its first widely
% % available version, mysteriously numbered 2.09, appeared in 1985.  \LaTeX{}
% % is now extremely popular in the scientific and academic communities, and
% % it is used extensively in industry.  It has become a \emph{lingua franca}
% % of the scientific world; scientists send their papers electronically to
% % colleagues around the world in the form of \LaTeX{} input.%
% % \end{munquote}

% The citation at the end is optional --- if you don't need it,
% then use \verb+\munquote+ without any arguments:

% \begin{munquote}%
% Here is a quote that does not have an associated citation
% after it.  You can specify the citation before or after the
% quote manually.%
% \end{munquote}

% By default, all text is double spaced, however, quotes and footnotes
% must be singled spaced.\munfootnote{This is a single spaced footnote.
% SGS requires that footnotes be singled spaced and this can be done with
% the \texttt{$\backslash$munfootnote} command.} The left margin is slightly
% wider than the right margin.  This is to compensate for binding.

% An example mathematical formulae is show in
% Equation~\ref{eqn:sum}.

% \begin{muneqn}{sum}
% \sum_{i = 0}^{n} i^2
% \end{muneqn}

% A slightly more complicated equation is given in Equation~\ref{eqn:schrodinger}:
% \munfootnote{Equation taken from the \textsl{Schr\"{o}dinger equation}
% entry on \textsl{Wikipedia}}

% \begin{muneqn}{schrodinger}
% i\hbar \frac{\partial}{\partial t}\Psi(x,\,t)=
% -\frac{\hbar^2}{2m}\nabla^2\Psi(x,\,t) + V(x)\Psi(x,\,t)
% \end{muneqn}

% % \section{Cross References}
% \label{sec:xrefs}

% In addition to using \verb+\ref+ to refer to equations, you can also use
% it (in conjunction with the \verb+\label+ command) to refer to sections
% and chapters without hard coding the numbers themselves.  For example,
% this is Section~\ref{sec:xrefs} of Chapter~\ref{chap:intro}.  You can
% also refer to Appendix~\ref{apdx:somelabel}, Subsection~\ref{sec:nested}
% below or any other place that has a \verb+\label+.  You can also use
% labels to refer to a page.  For example, Chapter~\ref{chap:figtab}
% starts on page~\pageref{chap:figtab}.

% % \section{Some Suggestions}

% Here are a few recommendations:

% \begin{itemize}
% 	\item Before using this template, make sure you check with
% 		your supervisor.
% 	\item MUN's library provides electronic access to some \LaTeX{}
% 		related textbooks which can be read online.  Use
% 		the search term \texttt{latex (computer file)} on the
% 		Library's web page.
% 	\item If you run into a problem, Google may be a helpful resource.
% 	\item Concentrate on content, let \LaTeX{} handle the typesetting.
% 	\item Don't worry about warnings related to:
% 	\begin{itemize}
% 		\item overfull \texttt{hboxes}/\texttt{boxes}
% 		\item underfull \texttt{hboxes}/\texttt{vboxes}
% 	\end{itemize}
% 	These can be corrected with modest rewording of your text prior
% 	to submission of your final copy.
% \end{itemize}

% % \section{The \texttt{Makefile}}

% You can use \texttt{make} to ``build'' your thesis on the Linux command
% line\munfootnote{Linux is available on all machines running LabNet in
% \textsl{The Commons} and in other computer labs on campus.} This will
% automatically run the \texttt{bibtex} program to create your bibliography
% and will also re-run \texttt{latex} as necessary to ensure that all
% references are resolved.  A device independent file (\texttt{thesis.dvi})
% will be created, by default.  If you are using this template in another
% environment other than the Linux command line, then the \texttt{Makefile}
% will probably not be useful to you.

% \begin{itemize}
% \item To make a PostScript copy of your thesis, type the following
% at the command line:

% \texttt{make thesis.ps}

% \item To generate a PDF copy of your thesis, run:

% \texttt{make thesis.pdf}

% \item To generate a PDF/A-1b copy of your thesis (which should
% satisfy the SGS's ethesis submission requirements):

% \texttt{make ethesis.pdf}

% \item To remove all the files generated by \texttt{bibtex} and
% \texttt{latex}, use the command:

% \texttt{make clean}

% \item To remove the intermediate files, but leave the PostScript
% and DVI/PDF files intact, use the command:

% \texttt{make neat}
% \end{itemize}

% As you add or remove figures, chapters, or appendices to your thesis,
% make sure you keep the \texttt{Makefile} upto date, too (see the
% \texttt{FIGURES} and \texttt{FILES} macros in the \texttt{Makefile}).

% % \section{Changing Fonts}

% Change fonts: {\Large Large},
% \verb+verbatim ~@#$%^&*(){}[]+,
% \textsc{Small Caps},
% \textsl{slanted text},
% \emph{emphasized text},
% \texttt{typewriter text}.

% % \section{Accents and Ligatures}

% Some accents:
% \'{e}
% \`{e}
% \^{o}
% \"{u}
% \c{c}
% \"{\i}
% \'{\i}
% \~{n}
% \={a}
% \v{a}
% \u{a}

% \noindent Some ligatures:
% fl{\ae}ffi
